PRACTICAL MACHINE LEARNING - Course Project
===========================================
*"How did they do the exercise ?"*

by Javier Martínez Arribas

## Loading and preprocessing the data
We can get the data from:

[Training Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)
[Testing data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The goal of this analysis is to predict the manner in which the subjects did the 
recorded exercise. For this purpose we have got two data sets, one training set 
and one testing set to apply our machine learning algorithm to twenty test cases.

We will split the training set into a train and a test set to make our model.
First of all we get the files from the web:

```{r}
library(caret)
#Training Set
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dest <- getwd()
dest <- paste(dest,sep="","/pml-training.csv")
download.file(url,destfile=dest,method="curl")
training <- read.csv(dest,header=T)

#Test Set
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
dest <- getwd()
dest <- paste(dest,sep="","/pml-testing.csv")
download.file(url,destfile=dest,method="curl")
testing <- read.csv(dest,header=T)
``` 

**Pre processing**

In order to address the problem of which predictors are better to construct the 
model we are going to make a deconstruction of the training set: 
First we separate the outcome from predictors. Outcome is the last column(classe):
```{r}
Y <- training[,160]
X <- training[,-160]
```

Then we can see what classes of predictors there are:
```{r}
table(sapply(X[1, ], class))
```

Once we have the types, we could divide the predictors set in order to explore 
deeply their importance for the final model.
```{r}
int <- sapply(X[1, ], class)=="integer"
X_int <- X[,int]
num <- sapply(X[1, ], class)=="numeric"
X_num <- X[,num]
fact <- sapply(X[1, ], class)=="factor"
X_fact <- X[,fact]
```
 
For the first subset formed by integer predictors we can considerate to remove 
some of them like:
```{r}
str(X_int)
```
The first predictor is just a counter of the rows in data set, so we can remove it:
```{r}
head(X_int$X)
tail(X_int$X)
```
Now, we can examinate the predictors with too much NAs at first glance:
```{r}
dim(X_int[complete.cases(X_int$max_picth_belt),])[1]
dim(X_int[complete.cases(X_int$min_pitch_belt),])[1]
dim(X_int[complete.cases(X_int$amplitude_pitch_belt),])[1]
dim(X_int[complete.cases(X_int$max_yaw_arm),])[1]
dim(X_int[complete.cases(X_int$min_yaw_arm),])[1]
dim(X_int[complete.cases(X_int$amplitude_yaw_arm),])[1]
dim(X_int)[1]
```
We discover that there are 406 complete cases out of 19622 in these predictors,
almost the 4%, so we could assume that there is not much relevant information into 
so we remove them as well.
We can remove also the time_stamp variables because we need to predict the way 
the exercise is done and there is no need to know the execution time.
```{r}
X_int <- subset(X_int,select=-c(X,raw_timestamp_part_1,raw_timestamp_part_2,num_window,
                                max_picth_belt,min_pitch_belt,amplitude_pitch_belt
                                      ,max_yaw_arm,min_yaw_arm,amplitude_yaw_arm))
dim(X_int[complete.cases(X_int),])[1]
```
As we can see now we have no NA values in this predictor subset.
We could study the correlation between predictors to avoid some redundant
information. We use the findCorrelation function in caret package.
```{r}
descrCor_int <- cor(X_int)
highlyCorDescr_int <- findCorrelation(descrCor_int, cutoff = 0.8)
X_int <- X_int[,-highlyCorDescr_int]
```
For the second predictors subset we are going to study numeric variables in the same way 
as integer ones:
```{r}
str(X_num)
```
Now we have got the same problem as with the integer variables there are some of
them with just 406 valid samples out of 19622, so we should remove them.
```{r}
for (i in 1:ncol(X_num)){
  if (i==1){nAs=NULL}
  nAs <- c(nAs,dim(X_num[complete.cases(X_num[,i]),])[1])
}
table(nAs)
```
and to remove just that columns...
```{r}
for (i in 1:ncol(X_num)){
  if (i==1){nam <- NULL}
  if ((dim(X_num[complete.cases(X_num[,i]),])[1]) == 406) {
    nam <- c(nam,colnames(X_num)[i])
  }
}
indices <- which(names(X_num) %in% nam)
X_num <- X_num[,-indices]
```
We can study the correlation between predictors in this subset as well:
```{r}
descrCor_num <- cor(X_num)
highlyCorDescr_num <- findCorrelation(descrCor_num, cutoff = 0.75)
X_num <- X_num[,-highlyCorDescr_num]
```

And the last step in order to get the most appropriate information for our model,
we study the most relevant factor predictors.
In this subset we have different cases to address to, like character variables in
factor format as well as date variables. We have also some factor variables with 
two levels, one of them "" and the other one "#DIV/0!", a kind of notation for 
error values in excel files. 
We have some numeric variables format as factor, categorical and binary ones.
I consider that we should not take into account this subset of predictors due to 
the few information contained into.
Just in order to show the huge number of zero values we are going to transform 
factor variables into numeric ones after removing some of no sense predictors.
```{r}
X_fact <- subset(X_fact,select=-c(new_window,user_name,cvtd_timestamp,kurtosis_yaw_belt,
                                         skewness_yaw_belt,amplitude_yaw_belt,
                                         kurtosis_yaw_dumbbell,skewness_yaw_dumbbell,
                                         amplitude_yaw_dumbbell,kurtosis_yaw_forearm,
                                         skewness_yaw_forearm,amplitude_yaw_forearm))
```
and transform the rest in numeric variables:
```{r warning=FALSE,message=FALSE}
for(i in 1:ncol(X_fact)){
  #Take the index where there are negative values
  ind <- grep("-",X_fact[,i])
  X_fact[,i] <- gsub("-","",X_fact[,i])
  X_fact[,i] <- invisible(as.numeric(X_fact[,i]))
  bad <- is.na(X_fact[,i])
  X_fact[bad,i] <- 0
  X_fact[ind,i]=X_fact[ind,i]*(-1)
}
head(X_fact)
```
We can see that a large number of rows are zero values.
```{r}
row_sub = apply(X_fact, 1, function(row) all(row ==0 ))
table(row_sub)
```
There are just 406 no zero rows out of 12622 rows. Almost a 4%
of the entire dataset.
After studying the different types of predictors we concatenate the predictors 
subsets together. Discarding factor transformed predictors:
```{r}
training_final <- cbind(Y,X_int,X_num)
```

Now we can define the control structure for training:
```{r}
set.seed(123)
ind <- createFolds(Y,returnTrain=TRUE)
cont <- trainControl(method="cv",index=ind)
```
For tune the desired model and calculate the accuracy of it we split the 
training set into a train and a test subset.
We fit a Random Forest method for this classification problem. In order not to waste
too much time nor computation resources we will select just ten trees and even so
we will get good predictions for our final exercise.
I did try to compute it with a higher number of trees in random forest but final 
predictions did not change.
```{r}
set.seed(123)
mtryVals <- floor(seq(10,ncol(training_final[,-1]),length=10))
mtryGrid <- data.frame(.mtry=mtryVals)
inTrain <- createDataPartition(training_final$Y,p = .8, list = FALSE)
train <- training_final[inTrain,]
test <- training_final[-inTrain,]
rfTune <- train(x = training_final[,-1], y = training_final[,1], method = "rf", 
                tuneGrid=mtryGrid,ntree=10, importance=TRUE, trControl = cont)
pred_train <- predict(rfTune,train)
ISerror <- sum(pred_train != train$Y) * 100 / nrow(train)
ISerror
```
We get an in sample error of almost 0.02%.
and we observe an accuracy and an out of sample error of:
```{r}
pred <- predict(rfTune,test)
table(pred,test$Y)
confusionMatrix(pred,test$Y)
OOSerror <- sum(pred != test$Y) * 100 / nrow(test)
OOSerror
```
We obtain a low out of sample error 0.05% and an accuracy of 99.95%.
Finally, we are going to use the testing file to complete the prediction exercise.
First we need to select just the predictors that we are going to need..
```{r}
indices <- which(names(testing) %in% names(train))
testing <- testing[,indices]
pred_final <- predict(rfTune,testing)
pred_final
```
Finally we get a random forest model for 43 predictors with a high accuracy and
a low out of sample error.
